### 2020-09-30 latex/vnhcs.tex
Started updating VNHCs as per Manfredi's comments
This is the beginning of updating the VNHC chapter according to
Manfredi's comments. I've moved some proofs around and started adding
punctuation to equations, but there is still much to do. Refer to the
draft_MMComments.pdf that I've annotated.

For the robotic systems of interest in this thesis, we assume that
each generalized coordinate \(q_i\) represents either the distance or the angle
between two parts of the system.
Mathematically, each \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to collect the coordinates into a \textit{configuration} 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system is a
so-called \textit{generalized cylinder}:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity} of the system. For arbitrary systems,
the space of allowable velocities depends on the current configuration of the
system.  However, since \(\mathcal{Q}\) is a generalized cylinder, we find that 
\(\dot{q} \in \R^n\).
The combined vector \((q,\dot{q}) \in \mathcal{Q}\times\R^n\) is called a 
\textit{state} of the system.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
\(\mathcal{L} : \mathcal{Q}\times\R^n \rightarrow \R\) defined by
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}.
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
When the mechanical system is actuated, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} (\ref{eqn:el-eqns-general}) with \textit{generalized input forces} 
\(\tau \in \R^k\) . 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i\tpose(q) \tau
\end{equation}
The vector \(B_i\tpose: \mathcal{Q} \rightarrow \R^{1\times k}\) describes how
the input forces shape the dynamics of \(q_i\).
The matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = \begin{bmatrix}
        - & B_1\tpose(q) & - \\
          & \vdots & \\
        - & B_n\tpose(q) & - \\
    \end{bmatrix}
\]
is called the \textit{input matrix} for the system.
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \((n - k)\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
%If this is the case, the Euler-Lagrange equations reduce to (\ref{eqn:el-eqns}),
%\begin{equation}\label{eqn:el-eqns}
%    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
%\end{equation}
%where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
%\[
%    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
%    \left(\pdiff{D_{i,j}}{q_k}  +
%     \pdiff{D_{i,k}}{q_j} -
%     \pdiff{D_{k,j}}{q_i}\right)\dot{q}_k
%\]

@@

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) := \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we write \(p = (p_1, \ldots, p_n) \in \R^n\) and call
\(p\) the \textit{conjugate of momenta to \(q\)}.
Note that each \(p_i\) is a linear function of \(\dot{q}\), and one can typically
solve for \(\dot{q}(q,p)\) by inverting all the expressions from (\ref{eqn:p-i}).
The combined vector \((q,p) \in \mathcal{Q}\times\R^n\) is called a 
\textit{phase} of the system.

The \textit{Hamiltonian} of the system in
\(\{q,p\}\) coordinates is the ``Legendre transform"
(\ref{eqn:hamiltonian-legendre}) of the Lagrangian \cite{landau_mechanics}.
\begin{equation}\label{eqn:hamiltonian-legendre}
    \mathcal{H}(q,p) := p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in the Hamiltonian framework are the \(2n\)
first-order equations called \textit{Hamilton's equations}. They are given by:
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
Here, \(B(q) \in \R^{n\times k}\) is the same input matrix used by the
Lagrangian framework, with \(\tau \in \R^k\) the same vector of generalized
input forces.

If the kinetic energy of the system is quadratic as in (\ref{eqn:lagrangian}), 
the conjugate of momenta becomes \(p = D(q)\dot{q}\). Since \(D(q)\) is symmetric
and positive definite, it is invertible at each \(q \in \mathcal{Q}\).
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

Any set of coordinates \(\{q,p\}\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}\) are
said to be \textit{canonical coordinates} for the system. A change of
coordinates \((q,p) \rightarrow (Q,P)\) is a \textit{canonical
transformation} if \(\{Q,P\}\) preserve the Hamiltonian structure; that is, if
they are canonical coordinates under the Hamiltonian
\(\mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

\section{Simply Actuated Hamiltonian Systems}
\textbf{Manfredi: What term should we use instead of ``simply actuated"?}

% Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Suppose we are given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}).
Because \(\tau\) is transformed by the input matrix \(B(q)\) before
entering the EOM, it is not in general clear how any particular input force \(\tau_i\)
will affect the dynamics of the system. 
In this section, we define a new class of Hamiltonian systems where the effect
of the input forces is made obvious. This class of
systems will form the backbone for the rest of the theory developed in this
thesis.

\begin{defn}
    Let \(\mathcal{H}\) be an \(n\)-DOF Hamiltonian system 
    with \(k \leq n\) actuators. 
    A set of canonical coordinates \(\{q,p\}\) for this system
    are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
    \]
    The first \((n-k)\) coordinates, labelled \(q_u\), are called the
    \textit{unactuated coordinates}. The remaining \(k\) coordinates, labelled
    \(q_a\), are called the \textit{actuated coordinates}. When grouping them
    together, we will always put them in the order \((q_u, q_a)\) to fit with 
    the definition. 
    The corresponding \((p_u, p_a)\) are called the unactuated and actuated
    momenta, respectively.
\end{defn}
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation from any canonical coordinates 
    \(\{q,p\}\) into simply actuated coordinates.
\end{defn}

@@

Note that if \(k = (n-1)\), the existence of any left annihilator 
\(A^0 \in \R^{1\times n}\) implies the left annihilator 
\(B^\perp := A^0/\norm{A^0}\) satisfies Assumption \ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    Assume \textbf{without loss of generality} that the input matrix \(B\) is
    left semi-orthogonal. That is, \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 \\
        0 & \sigma_2 & \cdots & 0 \\
        \vdots & & \ddots & \vdots \\
        0 & 0 & \cdots & \sigma_k \\
        - &   & \Zmat{(n-k)\times k} & -  \\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\) because \(B\) is full-rank \cite{calculating_svd}.
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau} \in \R^k\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal:
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
\]
Hence, \(\mathbf{B}\) is invertible with \(\mathbf{B}\inv = \mathbf{B}\tpose\).

@@

\begin{thm}\label{thm:simply-actuated}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \mathcal{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (\Id{n} \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &:= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\vspace{-4.5em} % For some reason not having this puts a large space before proof
\begin{proof}
    The Poisson bracket between the functions \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P_i\) and \(Q_j\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

    Therefore, by (45.10) in \cite{landau_mechanics}, the coordinate change 
    \((Q = \mathbf{B}q, P = \mathbf{B}p)\) is a canonical transformation.
    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the coordinates \(\{Q = (q_u,q_a), P = (p_u,p_a)\}\) are simply actuated coordinates for 
    \(\mathcal{H}\) as desired.
\end{proof}

>>>>

For the robotic systems of interest in this thesis, we assume that
each generalized coordinate \(q_i\) represents either the distance or the angle
between two parts of the system.
Mathematically, each \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to collect the coordinates into a \textit{configuration} 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system is a
so-called \textit{generalized cylinder}:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
    .
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity} of the system. For arbitrary systems,
the space of allowable velocities depends on the current configuration of the
system.  However, since \(\mathcal{Q}\) is a generalized cylinder, we find that 
\(\dot{q} \in \R^n\).
The combined vector \((q,\dot{q}) \in \mathcal{Q}\times\R^n\) is called a 
\textit{state} of the system.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
\(\mathcal{L} : \mathcal{Q}\times\R^n \rightarrow \R\) defined by
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}:
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
    .
\end{equation}
When the mechanical system is actuated, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} (\ref{eqn:el-eqns-general}) with \textit{generalized input forces} 
\(\tau \in \R^k\)
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i\tpose(q) \tau
    .
\end{equation}
The vector \(B_i\tpose: \mathcal{Q} \rightarrow \R^{1\times k}\) describes how
the input forces shape the dynamics of \(q_i\).
The matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = \begin{bmatrix}
        - & B_1\tpose(q) & - \\
          & \vdots & \\
        - & B_n\tpose(q) & - \\
    \end{bmatrix}
    ,
\]
is called the \textit{input matrix} for the system.
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \((n - k)\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
    ,
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
%If this is the case, the Euler-Lagrange equations reduce to (\ref{eqn:el-eqns}),
%\begin{equation}\label{eqn:el-eqns}
%    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
%\end{equation}
%where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
%\[
%    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
%    \left(\pdiff{D_{i,j}}{q_k}  +
%     \pdiff{D_{i,k}}{q_j} -
%     \pdiff{D_{k,j}}{q_i}\right)\dot{q}_k
%\]

@@

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) := \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
    .
\end{equation}
To ease notation, we write \(p = (p_1, \ldots, p_n) \in \R^n\) and call
\(p\) the \textit{conjugate of momenta to \(q\)}.
Note that each \(p_i\) is a linear function of \(\dot{q}\), and one can typically
solve for \(\dot{q}(q,p)\) by inverting all the expressions from (\ref{eqn:p-i}).
The combined vector \((q,p) \in \mathcal{Q}\times\R^n\) is called a 
\textit{phase} of the system.

The \textit{Hamiltonian} of the system in
\(\{q,p\}\) coordinates is the ``Legendre transform"
(\ref{eqn:hamiltonian-legendre}) of the Lagrangian \cite{landau_mechanics},
\begin{equation}\label{eqn:hamiltonian-legendre}
    \mathcal{H}(q,p) := p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
    .
\end{equation}
The EOM in the Hamiltonian framework are the \(2n\)
first-order equations called \textit{Hamilton's equations}. They are given by
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau
        .
    \end{cases}
\end{equation}
Here, \(B(q) \in \R^{n\times k}\) is the same input matrix used by the
Lagrangian framework, with \(\tau \in \R^k\) the same vector of generalized
input forces.

If the kinetic energy of the system is quadratic as in (\ref{eqn:lagrangian}), 
the conjugate of momenta becomes \(p = D(q)\dot{q}\). Since \(D(q)\) is symmetric
and positive definite, it is invertible at each \(q \in \mathcal{Q}\).
The Legendre transform (\ref{eqn:hamiltonian-legendre}) can be explicitly
written as
\[
    \mathcal{H}(q,p) = p\tpose D\inv(q)p - \frac{1}{2}p\tpose D\inv(q) p + V(q)
    ,
\]
which reduces to (\ref{eqn:hamiltonian}). 
Finding the derivative of each conjugate of momentum yields
\[
    \dot{p}_i = -\frac{1}{2} p\tpose \nabla_{q_i}D\inv(q) p 
        - \pdiff{P}{q_i}(q) + B_i\tpose(q) \tau
    ,
\]
which can be collected into the vector form (\ref{eqn:hamiltonian-full-dynamics}) 
by using the matrix Kronecker product.
In sum, the Hamiltonian system reduces to
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q)
    , \\
     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau
        . \\
    \end{cases} \label{eqn:hamiltonian-full-dynamics}
\end{align}

Any set of coordinates \((q,p)\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}\) are
said to be \textit{canonical coordinates} for the system. A change of
coordinates \((q,p) \rightarrow (Q,P)\) is a \textit{canonical
transformation} if \((Q,P)\) preserve the Hamiltonian structure; that is, if
they are canonical coordinates under the Hamiltonian
\(\mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

\citet{landau_mechanics} provides a useful result for showing whether a given
change of coordinates \((q,p) \to (Q,P)\) is a canonical transformation.
\begin{defn}
    The \textit{Poisson bracket} between the functions \(f(q,p)\) and \(g(q,p)\)
    is defined as
    \begin{equation}\label{eqn:poisson-bracket}
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
            \pdiff{f}{q_i}\pdiff{g}{p_i}
        .
    \end{equation}
\end{defn}
\begin{thm}\label{thm:canonical-transformations}
    A change of coordinates 
    \((q,p) \to (Q,P)\) is a canonical transformation if
    \begin{align*}
        [Q_i, Q_j] &= 0
        , \\
        [P_i, P_j] &= 0
        , \\
        [P_i, Q_j] &= \delta_{i,j}
        , 
    \end{align*}
    for all \(i,j \in \n\).
\end{thm}
\begin{proof}
    See (45.10) in \cite{landau_mechanics}.
\end{proof}

Later in this chapter we will define a particular type of coordinate change.
The following Lemma allows us to prove it is a canonical transformation.

\begin{lemma}\label{lemma:linear-is-canonical}
    Let \(\mathcal{H}\) be a Hamiltonian system in canonical coordinates
    \((q,p)\).
    Let \(A \in \R^{n\times n}\) be an invertible matrix such that
    \(A\inv = A\tpose\).
    The change of coordinates \((q,p) \to (Q = Aq, P = Ap)\) is a canonical
    transformation.
\end{lemma}
\begin{proof}
    For any constant matrix \(A\), the transformation
    \((Q = Aq, P = Ap)\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, 
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0
        , \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
        .
    \end{align*}
    Note that \((A_i)\tpose (A\tpose)_j = (A_i)\tpose (A\inv)_j = \delta_{i,j}\). 
    Using this fact we see that the Poisson brackets between \(P_i\) and \(Q_j\)
    are given by
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n A_{i,m}A_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n A_{i,m}A\tpose_{m,j} \\
                  &= (A_i)\tpose A\tpose_j \\
                  &= \delta_{i,j}
        .
    \end{align*}
    Therefore, by Theorem \ref{thm:canonical-transformations}, the coordinate
    change \((Q = Aq, P = Ap)\) is a canonical transformation.
\end{proof}

\section{Simply Actuated Hamiltonian Systems}
% Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Suppose we are given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}).
Because \(\tau\) is transformed by the input matrix \(B(q)\) before
entering the EOM, it is not in general clear how any particular input force \(\tau_i\)
will affect the dynamics of the system. 
In this section, we define a new class of Hamiltonian systems where the effect
of the input forces is made obvious. This class of
systems will form the backbone for the rest of the theory developed in this
thesis.

\begin{defn}
    Let \(\mathcal{H}\) be an \(n\)-DOF Hamiltonian system 
    with \(k \leq n\) actuators. 
    A set of canonical coordinates \((q,p)\) for this system
    are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
        .
    \]
    The first \((n-k)\) coordinates, labelled \(q_u\), are called the
    \textit{unactuated coordinates}. The remaining \(k\) coordinates, labelled
    \(q_a\), are called the \textit{actuated coordinates}. When grouping them
    together, we will always put them in the order \((q_u, q_a)\) to fit with 
    the definition. 
    The corresponding \((p_u, p_a)\) are called the \textit{unactuated} and 
    \textit{actuated momenta}, respectively.
\end{defn}
% TODO: delete this next definition and rephrase the text accordingly
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation from any canonical coordinates 
    \(\{q,p\}\) into simply actuated coordinates.
\end{defn}

@@

Assumption \ref{assm:B-perp} requires the rows of \(B^\perp\) to be unit vectors
that are mutually orthogonal. 
In the case that \(k = (n-1)\), the existence of any left annihilator 
\(A^0 \in \R^{1\times n}\) implies the left annihilator 
\(B^\perp := A^0/\norm{A^0}\) will satisfy Assumption \ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    Assume \textbf{without loss of generality} that the input matrix \(B\) is
    left semi-orthogonal. That is, \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 \\
        0 & \sigma_2 & \cdots & 0 \\
        \vdots & & \ddots & \vdots \\
        0 & 0 & \cdots & \sigma_k \\
        - &   & \Zmat{(n-k)\times k} & -  \\
    \end{bmatrix}
    ,
\]
where \(\sigma_i \neq 0\) because \(B\) is full-rank \cite{calculating_svd}.
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
    ,
\]
and defining the regular feedback transformation\(\tau = V T \hat{\tau}\), 
we get a new input matrix for \(\hat{\tau} \in \R^k\) given by 
\[
    \hat{B} = B V T = 
    U \begin{bmatrix}
        \Id{k} \\ \Zmat{(n-k)\times k}
    \end{bmatrix}
    ,
\]
which is still constant and full-rank. 
In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
    .
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal:
\begin{proof}
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
    .
\]
Hence, \(\mathbf{B}\) is invertible with \(\mathbf{B}\inv = \mathbf{B}\tpose\).
\end{proof}

@@

\begin{thm}\label{thm:simply-actuated}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \mathcal{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q)
        , \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (\Id{n} \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose
        , \\
        V(Q) &:= P(\mathbf{B}\tpose Q)
        .
    \end{align*}
\end{thm}
\vspace{-4.5em} % For some reason not having this puts a large space before proof
\begin{proof}
    The change of coordinates \((Q = \mathbf{B}q, P = \mathbf{B}p)\) is a
    canonical transformation because it satisfies Lemma
    \ref{lemma:linear-is-canonical}.
    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
        ,
    \]
    so \(\left(Q = (q_u,q_a), P = (p_u,p_a)\right)\) are simply actuated
    coordinates for \(\mathcal{H}\) as desired.
\end{proof}
